\documentclass[12pt]{letter}
\usepackage{geometry}
\geometry{margin=2.5cm}
\usepackage{hyperref}

\signature{Livija Jakaite and Vitaly Schetinin\\
  School of Computing and Engineering\\
  University of Bedfordshire, UK\\
  \texttt{Vitaly.Schetinin@beds.ac.uk} (corresponding)}

\address{%
  Livija Jakaite and Vitaly Schetinin\\
  School of Computing and Engineering\\
  University of Bedfordshire\\
  Luton, UK\\[0.5em]
  \today
}

\begin{document}
\begin{letter}{%
  Editor-in-Chief\\
  \textit{Bayesian Analysis}\\
  International Society for Bayesian Analysis}

\opening{Dear Editor,}

We are pleased to submit the manuscript

\begin{quote}
\textbf{``Catalan-Exponential Priors for Bayesian Decision Trees:
WAIC Consistency, Posterior Concentration, and PAC-Bayes Guarantees''}
\end{quote}

for consideration for publication in \textit{Bayesian Analysis}.

\medskip\noindent\textbf{Summary of contributions.}\quad
This paper develops the theoretical foundations of the JBDT model
(Jakaite \& Schetinin, 2025, MDPI Make), a Bayesian decision tree classifier
that uses a Catalan-exponential tree-size prior and Dirichlet-Multinomial
leaf models with RJMCMC inference.
We prove four main results that collectively provide a complete theoretical
justification for the model design and explain previously unexplained empirical findings:

\begin{enumerate}
\item \textbf{WAIC-LOO gap theorem}: The WAIC and LOO deviances agree
  up to the constant $-4N/(N+\alpha_0)+O(N^{-1})$, proved in closed form.
  This yields the practical design criterion $N_{\min} \approx 5.41/\Delta$,
  which predicts the exact sample size at which WAIC-weighted BMA
  becomes reliable.

\item \textbf{Catalan prior tail bound}: $P(k \geq k_0) \leq C_\gamma\,(e^{-\gamma}/4)^{k_0}$,
  with universal effective decay rate $r_{\rm eff} = 1.228\,e^{-\gamma}/4$.
  Posterior concentration on the true complexity $k^*$ is achieved at $N=300$
  (Catalan) versus $N>800$ (Chipman 1998).

\item \textbf{BMA oracle inequality}: $R_{\rm BMA}-R_{k^*} \leq \Delta_{\rm Jensen}
  = O(\exp(-N\Delta_{\rm WAIC}/2))$.
  Excess risk decreases 56-fold from $N=20$ to $N=800$, following $O(1/N)$.

\item \textbf{PAC-Bayes sample complexity}: Under the oracle posterior,
  the Catalan prior requires $8.1\times$ fewer samples than Chipman for
  sparse true models ($k^*=1$), directly explaining its advantage in
  small-$N$ medical imaging.
\end{enumerate}

Supplementary appendices provide RJMCMC detailed balance, ECE calibration bounds,
and decision-theoretic analysis connecting the above to asymmetric-loss classification.
All results are accompanied by numerical verification on both synthetic and
real knee osteoarthritis data.

\medskip\noindent\textbf{Fit for Bayesian Analysis.}\quad
The paper is squarely within the scope of \textit{Bayesian Analysis}:
it develops methodology for Bayesian model selection (WAIC) and BMA,
analyses a novel prior construction (Catalan-exponential) analytically,
and provides PAC-Bayesian guarantees.
The application to medical imaging provides concrete motivation but the
theoretical results are model-agnostic within the DM-leaf class.

\medskip\noindent\textbf{Suggested associate editors / reviewers.}\quad
Given the combination of Bayesian nonparametric trees, model selection,
and PAC-Bayes bounds, We would suggest reviewers with expertise in
Bayesian model selection, decision trees, or PAC-Bayesian theory.

\medskip\noindent\textbf{Declaration.}\quad
The manuscript is original, has not been published, and is not under
consideration elsewhere.
There are no conflicts of interest.
The paper is \textbf{not anonymised} in this version; please advise if
double-blind anonymisation is required for the current submission round.

\medskip\noindent\textbf{Code availability.}\quad
All experiments are reproducible; code will be made available at
\url{https://github.com/vitsch/jbdt} upon acceptance.

\closing{Yours sincerely,}

\end{letter}
\end{document}
