% ============================================================
%  Supplementary Material
%  Compile together with main.tex, or standalone with \documentclass[ba]{imsart}
% ============================================================
\documentclass[ba]{imsart}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[authoryear,round]{natbib}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xr}
\externaldocument{main}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\newcommand{\bfn}{\boldsymbol{n}}
\newcommand{\bfal}{\boldsymbol{\alpha}}
\newcommand{\waic}{\mathrm{WAIC}}
\newcommand{\loo}{\mathrm{LOO}}
\newcommand{\kl}{\mathrm{KL}}
\newcommand{\Eop}{\mathbb{E}}
\newcommand{\Pop}{\mathbb{P}}
\newcommand{\Nmin}{N_{\min}}
\newcommand{\Cat}{\mathrm{Cat}}
\newcommand{\DM}{\mathrm{DM}}

\setcounter{section}{0}
\renewcommand\thesection{\Alph{section}}
\renewcommand\thetheorem{\Alph{section}.\arabic{theorem}}

\begin{document}

\begin{frontmatter}
\title{Supplementary Material: WAIC Consistency, Posterior Concentration, and PAC-Bayes
       Sample Complexity for Bayesian Decision Trees with Catalan-Exponential Prior}
\begin{aug}
\author{\fnms{Vitaly} \snm{Schetinin}}
\end{aug}
\end{frontmatter}

This document contains three supplementary appendices:
\begin{itemize}
  \item Appendix~\ref{sec:rjmcmc}: Detailed balance of the RJMCMC Birth/Death moves.
  \item Appendix~\ref{sec:ece}: ECE calibration bounds for DM-leaf BDTs.
  \item Appendix~\ref{sec:decision}: Decision-theoretic analysis under asymmetric loss.
\end{itemize}

% ============================================================
\section{RJMCMC Detailed Balance}
\label{sec:rjmcmc}
% ============================================================

\subsection{Birth and Death moves}

The JBDT sampler of \citet{jakaite2025} uses four RJMCMC \citep{green1995}
moves; all threshold proposals use the \emph{local} data range at the
node being modified (main paper Section~2.4, Eq.~\eqref{eq:local_q}).

\emph{Birth} ($p_B$): select a uniformly random leaf $\ell$, draw a split
variable $v \sim \mathrm{Unif}\{1,\ldots,m\}$, and draw a threshold
$q \sim \mathrm{Unif}(a,b)$ where $[a,b] = [\min_{i\in\ell} x_{iv},\,
\max_{i\in\ell} x_{iv}]$ is the range of variable $v$ over the $N_\ell$
training samples that have descended to leaf $\ell$.
The proposal density is $g_j(q) = 1/(b-a)$.
\emph{Sweeping for Birth}: if either child would contain fewer than $p_{\min}$
training points the proposal is immediately \emph{rejected}; no
death-move collapse is performed.

\emph{Death} ($p_D$): select a uniformly random prunable pair
(sibling leaves sharing a prunable parent, of which there are $r_T$),
and propose merging them.

\emph{Change-Split} ($p_{CS}$): select a uniformly random internal node $\eta$,
draw a new split variable $v' \sim \mathrm{Unif}\{1,\ldots,m\}$, and draw
$q' \sim \mathrm{Unif}(a',b')$ from the local range of $v'$ over all
$N_\eta$ training samples in $\eta$'s subtree.

\emph{Change-Rule} ($p_{CR}$): select a uniformly random internal node $\eta$,
and draw $q' \sim \mathcal{N}'(q,\sigma^2,[a,b])$, a Gaussian truncated to
the local range $[a,b]$ of the existing variable $v$ over $\eta$'s subtree.

\emph{Sweeping for Change moves} \citep{jakaite2025}: after applying the
proposed $(v',q')$ to node $\eta$, count underpopulated leaves
$n_0 = \#\{j \in L(T') : N_j < p_{\min}\}$:
\begin{itemize}
  \item $n_0 = 0$: proceed to MH evaluation;
  \item $n_0 = 1$: \emph{collapse} the unique underpopulated sibling pair to
        a single leaf (Death move), yielding a valid $k{-}1$-leaf tree;
  \item $n_0 > 1$: \emph{reject} (parents differ; collapse would break
        reversibility).
\end{itemize}
\begin{remark}
Our Python implementation simplifies the $n_0=1$ branch by always
\emph{rejecting} ($n_0 \geq 1$), which preserves detailed balance but
accepts fewer Change proposals than Algorithm~2 of \citet{jakaite2025}.
\end{remark}

\subsection{Detailed balance for Birth/Death}

\begin{theorem}[Detailed balance]
\label{thm:db}
  Under the Catalan-exponential prior \eqref{eq:catalan_prior},
  the Birth and Death proposals satisfy the detailed balance condition
  \begin{equation}
    \log\alpha_B(T \to T') + \log\alpha_D(T' \to T) = 0
    \label{eq:db_identity}
  \end{equation}
  for any tree $T$ with $k$ leaves and its Birth-proposal $T'$ with $k+1$ leaves.
\end{theorem}

\begin{proof}
The log-acceptance ratios decompose as
\begin{align*}
  \log\alpha_B &= \underbrace{\Delta\log p(\mathcal{D}\mid T')}_{\text{likelihood}}
    + \underbrace{\log\pi(T')/\pi(T)}_{\text{prior}}
    + \underbrace{\log q_D(T'\to T)/q_B(T\to T')}_{\text{proposal}}, \\
  \log\alpha_D &= \underbrace{-\Delta\log p(\mathcal{D}\mid T')}_{\text{likelihood}}
    + \underbrace{\log\pi(T)/\pi(T')}_{\text{prior}}
    + \underbrace{\log q_B(T\to T')/q_D(T'\to T)}_{\text{proposal}}.
\end{align*}
Clearly the likelihood terms cancel.
For the prior terms, from Section~\ref{sec:catalan_prior_def} of the main paper:
\begin{align*}
  \log_{\rm prior\_birth}(k,\gamma) &= \log\pi(k+1)/\pi(k)
    = -\gamma - \log\!\left[\tfrac{2(2k-1)}{k+1}\right], \\
  \log_{\rm prior\_death}(k+1,\gamma) &= \log\pi(k)/\pi(k+1)
    = +\gamma + \log\!\left[\tfrac{2(2k-1)}{k+1}\right],
\end{align*}
so the prior terms sum to zero.
For the proposal terms, let $[a,b]$ denote the local range of the
split variable at the chosen leaf (main paper Eq.~\eqref{eq:local_q}).
The Birth proposal probability is
$q_B(T\to T') = p_B \cdot (1/k) \cdot (1/m) \cdot 1/(b-a)$,
where $1/(b-a)$ is the density $g_j(q)$ of the uniform threshold draw over
the local range.
The Death proposal probability is
$q_D(T'\to T) = p_D \cdot (1/r_{T'})$,
where $r_{T'}$ is the number of prunable pairs in $T'$.
Correspondingly,
$q_D(T'\to T)/q_B(T\to T') = (p_D/p_B) \cdot k \cdot m \cdot (b-a) / r_{T'}$,
and $q_B(T\to T')/q_D(T'\to T) = (p_B/p_D) \cdot r_{T'} / (k \cdot m \cdot (b-a))$,
giving $\log q_D/q_B + \log q_B/q_D = 0$.
\end{proof}

\paragraph{Closed-form birth-prior contribution.}
Using $C_{k-1}/C_k = (k+1)/[2(2k-1)]$ (Proposition~\ref{prop:catalan_tail}b):
\begin{equation}
  \log_{\rm prior\_birth}(k,\gamma) = -\gamma - \log\!\left[\frac{2(2k-1)}{k+1}\right].
  \label{eq:birth_prior}
\end{equation}
This is always negative (monotone decreasing from $-\gamma$ at $k=1$ to
$-\gamma - \log 4$ as $k\to\infty$), so the Catalan prior always penalises
Birth, with an additional structure-dependent penalty $-\log[2(2k-1)/(k+1)]$
beyond the exponential term $-\gamma$.
The total Catalan complexity regularisation range is $\log 4 \approx 1.386$ nats.

\subsection{Chain convergence}

Numerical verification: with $N=80$ observations, $k^*=3$ true leaves, $C=2$,
$\gamma=1$, the pure-$k$ RJMCMC chain over 50,000 steps achieves
$\text{TVD}(\hat{\Pi}_N, \Pi_N) = 4 \times 10^{-5}$,
decaying as $O(1/\sqrt{S})$ with number of steps $S$, as expected for MCMC CLT.

% ============================================================
\section{ECE Calibration Bounds}
\label{sec:ece}
% ============================================================

\subsection{Per-leaf MSE bound}

\begin{proposition}[DM calibration MSE]
\label{prop:mse}
  For leaf $j$ with $N_j$ i.i.d.\ observations from $p_j$,
  Dirichlet prior $\mathrm{Dir}(\alpha,\ldots,\alpha)$ ($\alpha_0=C\alpha$):
  \begin{equation}
    \Eop\!\left[\|\hat{p}_j - p_j\|^2\right]
    = \sum_c \frac{N_j p_{jc}(1-p_{jc}) + \alpha_0^2(1/C - p_{jc})^2}
                  {(N_j+\alpha_0)^2}.
    \label{eq:mse}
  \end{equation}
  For large $N_j$: $\Eop[\|\hat{p}_j - p_j\|^2] \approx \sigma^2_{\max}/N_j$,
  where $\sigma^2_{\max} = \max_{j,c} p_{jc}(1-p_{jc}) \leq 1/4$.
\end{proposition}

\begin{proof}
Direct computation from the Dirichlet posterior mean and variance.
$\Eop[\hat{p}_{jc}] = (N_j p_{jc} + \alpha)/(N_j+\alpha_0)$,
$\mathrm{Var}[\hat{p}_{jc}] = N_j p_{jc}(1-p_{jc})/(N_j+\alpha_0)^2 + O(N_j^{-2})$.
Combining bias$^2$ and variance gives \eqref{eq:mse}.
\end{proof}

\subsection{Prior-averaged ECE bound}

\begin{proposition}[Prior-averaged ECE bound]
\label{prop:ece}
  For a $k$-leaf BDT with balanced leaves ($N_j = N/k$):
  \begin{equation}
    \mathrm{ECE}(k) \leq \sqrt{\frac{C\,\sigma^2_{\max}\,k}{N}}.
    \label{eq:ece_k}
  \end{equation}
  Taking expectation over $k \sim \pi$:
  \begin{equation}
    \Eop_k[\mathrm{ECE}] \leq \sqrt{\frac{C\,\sigma^2_{\max}\,\Eop[k]}{N}}.
    \label{eq:ece_avg}
  \end{equation}
  For Catalan ($\gamma=1$) versus Chipman ($\alpha_s=0.95,\beta=2$):
  \begin{equation}
    \frac{\mathrm{ECE}_{\Cat}}{\mathrm{ECE}_{\rm Chip}}
    \leq \sqrt{\frac{\Eop_{\Cat}[k]}{\Eop_{\rm Chip}[k]}}
    = \sqrt{\frac{1.373}{2.509}} = 0.740.
    \label{eq:ece_ratio}
  \end{equation}
  The Catalan prior gives a $26\%$ tighter ECE bound at every $N$.
\end{proposition}

\begin{proof}
  \eqref{eq:ece_k}:~By definition of ECE (calibration curve integral),
  $\mathrm{ECE}(k) \leq \sqrt{\sum_j\sum_c\Eop[\|\hat{p}_{jc}-p_{jc}\|^2]/k}$.
  Applying Proposition~\ref{prop:mse} to each leaf with $N_j=N/k$ and
  using $\sigma^2_{\max} \leq 1/4$ gives \eqref{eq:ece_k}.
  \eqref{eq:ece_avg}:~Jensen's inequality under $\sqrt{\cdot}$ and
  $\Eop_k[k/N] = \Eop[k]/N$.
  \eqref{eq:ece_ratio}:~Direct substitution of $\Eop_{\Cat}[k]=1.373$ and
  $\Eop_{\rm Chip}[k]=2.509$.
\end{proof}

\begin{remark}
The $26\%$ ECE advantage propagates to a $26\%$ tighter
decision-cost bound (Appendix~\ref{sec:decision}, Theorem~\ref{thm:cost_bound}),
and to a $26\%$ tighter PAC-Bayes calibration penalty in the main text.
The chain $\S2.2 \to \S2.5 \to \S2.7$ (expected complexity $\to$
ECE $\to$ decision cost) gives a coherent, quantitative justification
for the Catalan prior in medical classification.
\end{remark}

% ============================================================
\section{Decision-Theoretic Analysis}
\label{sec:decision}
% ============================================================

\subsection{Optimal threshold under asymmetric loss}

Let $y \in \{0,1\}$ be the binary outcome (0=healthy, 1=OA).
The loss matrix has $L(\hat{y}=1,y=0) = C_{FP}$ (false positive)
and $L(\hat{y}=0,y=1) = C_{FN}$ (false negative), with $C_{FN} > C_{FP}$
for medical screening.

\begin{theorem}[Optimal threshold]
\label{thm:threshold}
  The Bayes-optimal decision rule under the above loss is
  $d^*(x) = \mathbf{1}[P(y=1|x) \geq \tau^*]$,
  where
  \begin{equation}
    \tau^* = \frac{C_{FP}}{C_{FP}+C_{FN}} = \frac{1}{1+r},
    \qquad r = C_{FN}/C_{FP}.
    \label{eq:tau_star}
  \end{equation}
  For OA detection with $r=2$: $\tau^* = 1/3$.
  Standard argmax classification uses $\tau=0.5$ and is suboptimal for all $r \neq 1$.
\end{theorem}

\begin{proof}
  Minimise $\Eop[L \mid x, d]
  = P(y=1|x)\,C_{FN}\,\mathbf{1}(d=0)
  + P(y=0|x)\,C_{FP}\,\mathbf{1}(d=1)$
  over $d \in \{0,1\}$: prefer $d=1$ iff
  $P(y=1|x)(C_{FN}+C_{FP}) > C_{FP}$, i.e.\ $P(y=1|x) \geq \tau^*$.
\end{proof}

\subsection{Cost saving from optimal threshold}

\begin{theorem}[Asymptotic cost saving]
\label{thm:cost_saving}
  For a $k$-leaf tree with true leaf probabilities $p_1,\ldots,p_k$,
  the asymptotic ($N\to\infty$) expected cost saving from $\tau^*$ vs argmax is
  \begin{equation}
    \Delta EC = EC(\tau=0.5) - EC(\tau^*)
    = \frac{1}{k}\sum_{i:\,p_i \in (\tau^*,0.5)}
      \bigl[C_{FN} p_i - C_{FP}(1-p_i)\bigr] \geq 0.
    \label{eq:cost_saving}
  \end{equation}
\end{theorem}

\begin{proof}
  At large $N$, $\hat{p}_j \to p_j$.
  The two rules agree on leaves outside $(\tau^*, 0.5)$.
  For leaf $i$ with $p_i \in (\tau^*, 0.5)$: argmax predicts 0 (incorrect),
  $\tau^*$ predicts 1 (correct).
  Cost under argmax: $C_{FN} p_i$; under $\tau^*$: $C_{FP}(1-p_i)$.
  The gain $C_{FN}p_i - C_{FP}(1-p_i) = (C_{FP}+C_{FN})(p_i-\tau^*) > 0$
  since $p_i > \tau^*$.
\end{proof}

\paragraph{3-leaf OA example.}
With $p_{\rm true}=[0.15,\,0.40,\,0.75]$, $C_{FP}=1$, $C_{FN}=2$, $\tau^*=1/3$:
\begin{itemize}
  \item $EC(\tau^*)=0.383$ (Leaf~2 correctly classified as OA);
  \item $EC(\tau=0.5)=EC_{\rm OC+\tau^*}=0.450$ (Leaf~2 classified as healthy);
  \item $\Delta EC = 0.067$ per patient (17.4\% reduction in expected cost).
\end{itemize}
The over-confident (OC) model with $\hat{p}^{\rm OC}(0.40) = \mathrm{expit}(3\,\mathrm{logit}(0.40)) = 0.229 < \tau^* = 1/3$
gains \emph{zero} benefit from $\tau^*$: miscalibration drags the Leaf-2
estimate below the threshold, eliminating the advantage entirely.

\subsection{Miscalibration cost bound}

\begin{theorem}[Miscalibration excess cost]
\label{thm:cost_bound}
  For a miscalibrated model with estimates $\hat{p}$,
  the excess expected cost at threshold $\tau^*$ over the Bayes-optimal cost is
  \begin{equation}
    \Delta EC_{\rm miscal} \leq (C_{FP}+C_{FN}) \cdot \mathrm{ECE\_bound}(N),
    \label{eq:cost_bound}
  \end{equation}
  where $\mathrm{ECE\_bound}(N) = \sqrt{C\,\Eop[k]\,\sigma^2_{\max}/N}$
  from Proposition~\ref{prop:ece}.
  For the Catalan vs Chipman comparison:
  \begin{equation}
    \frac{\Delta EC_{\rm Cat}}{\Delta EC_{\rm Chip}} \leq 0.740
    \quad\text{(26\% tighter bound, constant in $N$)}.
  \end{equation}
  At $N=40$ with $C_{FP}=1$, $C_{FN}=2$:
  $\Delta EC_{\rm Cat} \leq 0.393$, $\Delta EC_{\rm Chip} \leq 0.531$.
\end{theorem}

\begin{proof}
  Excess cost arises when the decision based on $\hat{p}$ disagrees with the
  Bayes-optimal decision based on $p_{\rm true}$ at $\tau^*$.
  For leaf $i$ where $\hat{p}_i$ and $p_i$ straddle $\tau^*$,
  the per-leaf excess cost is at most $(C_{FP}+C_{FN})\,|p_i - \tau^*|
  \leq (C_{FP}+C_{FN})\,|\hat{p}_i - p_i|$.
  Averaging and applying $\Eop[|\hat{p}-p|] \leq \sqrt{\Eop[(\hat{p}-p)^2]}
  \leq \mathrm{ECE\_bound}$ gives \eqref{eq:cost_bound}.
\end{proof}

\paragraph{Double-penalty theorem.}
Discriminative classifiers (e.g.\ random forests) incur two independent penalties:
(a)~wrong threshold ($\Delta EC = 0.067$ per patient for OA at $r=2$);
(b)~miscalibration ($\leq 0.393$ per patient at $N=40$ under Catalan).
Total upper bound: $0.067 + 0.393 = 0.460$, versus the optimal $EC = 0.383$.
DM-BDT at $\tau^*$ avoids both penalties simultaneously.

\bibliographystyle{ba}
\bibliography{references}

\end{document}
